Apparent preferences for cognitive effort fade when multiple forms of effort

### 2021-05-28 Manuscript.Rmd
revised intro

  Evaluating whether to pursue a desirable outcome often entails assessing the demands required for obtaining it. In psychology and economics, demands for both time and effort have traditionally been identified as costs that detract from the net subjective value of rewards [@Shenhav2017; @Westbrook2015; @Frederick2002; @Walton2006]. Indeed, decision makers tend to prefer low-effort and immediate rewards over larger rewards that demand longer delays and higher effort [@Ainslie1975; @Hull1943; @Kool2010; @Kool2018; @Treadway2009; @Westbrook2013]. However, people still needlessly pursue effortful courses of action in everyday decisions. Emerging experimental and theoretical work has chiefly explained this phenomenon by suggesting a reciprocal relationship between effort and outcomes, whereby exerting effort boosts the perceived value of its outcomes [@HernandezLallement2014; @Kacelnik2002; @Mochon2012], and positive outcomes imbue value into their preceding effortful behavior [i.e. learned industriousness, @Eisenberger1992]. Though intuitively valid and empirically supported, this explanation leaves open questions about what features of an effortful task can make it attractive in its own right, and how contextual factors mold its perceived costs [@Inzlicht2018]. Understanding what governs the intrinsic costs and rewards of effort could improve our ability to motivate completion of effortful daily activities, such as schoolwork or physical exercise.

  Recent work has proposed that the subjective costs produced by delays and cognitive effort are modulated by the possibility of missing out on better concurrent alternatives, implying that decision makers track the *opportunity cost of time* [OC, @Fawcett2012; @Kurzban2013; @Otto2019]. Preferences for effortful courses of action could therefore arise from the presence of other types of demands and opportunities in the same choice environment [@Kool2018]. For example, exerting high effort for a reward might be aversive, but the same level of work might be appealing in comparison to the boredom of an equivalent period of passive waiting (e.g. taking the cumbersome way home rather than waiting for a delayed train, even if the resulting time of arrival is the same). This effect of context on effort preferences has been illustrated by experiments on charitable giving, in a which high-effort donation method (such as a run) attracted larger donations than a low-effort method (such as a picnic) when either was available alone (the “martyrdom effect”), but the high-effort event was disfavored when both options were offered as alternatives [@Olivola2013].

  Standard models in neuroeconomics and behavioral economics cannot readily address this possibility. The effects on valuation of different types of effort or delay have traditionally been formalized with discounting functions that capture how much less appealing a reward becomes as a function of the associated amount of effort or delay [@Ainslie1975; @Frederick2002; @Green1994; @Kable2007; @kleinflugge2015; @Kool2010; @Westbrook2013]. An advantage of relying on these standardized functions is that they can circumvent a key confound when comparing effort and delay, namely that increasingly effortful tasks also tend to take longer to perform. However, although this line of work suggests that discounting functions are similar for delay and effort [@Massar2015; @Prevost2010; @Seaman2018], there is ongoing debate about the shape of the discounting function across types of demands [@Arulpragasam2018; @Biaaszek2017; @Chong2017; @Hartmann2013; @Klein-Flugge2016; @Kool2018; @Prevost2010], and the degree to which discount rates correlate across cost domains [@Seaman2018; @Westbrook2013]. Further, this approach does not directly probe which type of demand people would prefer if given the option, and constrains the understanding of behavioral demands to one in which such demands are bound to reduce the value of prospects.

  This evidence sets the stage for examining what people would actually select in environments that intermixed multiple types of demands, including cognitive, physical effort, and delay. Single choice foraging paradigms provide a natural way to interrogate preferences among demands, by asking individuals whether the current offer outweighs what could be obtained and experienced during the same time period instead [@Constantino2015; @Garrett2020; @Krebs1977; @Mobbs2018; @Stephens1986]. Such paradigms probe the perceived opportunity cost of time, which has been hypothesized to underlie the evaluation of behavioral costs [@Kurzban2013]. Specifically, foragers must decide whether to spend their limited time by harvesting their currently available resource or by seeking out more profitable (or less costly) alternatives. This could involve, for example, deciding when to leave a patch of depleting richness (e.g. when to quit a job that provides diminishing opportunities for growth), or choosing whether to pursue or forgo offers presented serially (e.g. whether to accept your first job offer or wait for new potential ones). Here we focus on this latter type of foraging, traditionally called prey foraging [@Krebs1977], as it provides a means to examine how choices to accept or reject individual prospects are influenced by contextual knowledge about other potentially obtainable alternatives in the same environment.

  Foraging has recently attracted high scientific interest due to its ecological validity, which is rooted in evolutionarily conserved choice behaviors [@Hayden2018; @Mobbs2018], and offers a number of desirable qualities to address our question. First, it allows for the estimation of optimal, reward maximizing choice patterns on the basis of the opportunity cost of time [@Charnov1976; @Krebs1977]. The opportunity cost depends on the richness of the environment, which can be experimentally manipulated by balancing the amount of time it takes to harvest an offer (*handling time*) versus to search for new ones (*travel time*). We can then interpret observed deviations from such optimal behavior, and examine how foraging behavior changes when the handling time is filled with cognitive effort or physical effort in comparison with unfilled delay (thus disentangling effort from delay duration). Second, in contrast to multi-alternative economic choice paradigms that often treat choices as independent observations, the sequential nature of foraging encourages individuals to consider both focal and global elements of the decision context. For example, the cost (or value) of effort and delay could globally depend on the availability of alternative demands, impacting the perceived opportunity cost of time. Alternatively, it could be intrinsic and robust to such manipulation, and/or depend on task-specific factors that are adjusted once individuals experience different forms of behavioral costs. Foraging rests on an abundant modeling history that provides computational tools to formalize relevant decision variables [@Stephens1986].

  Across two behavioral experiments, we tested the overarching hypothesis that preferences would vary depending on whether individuals faced a single type or multiple types of behavioral demands (i.e. cognitive effort, physical effort, and delay). Experiment 1 used a between-subject design in which participants foraged for rewards of varying magnitudes that imposed varying durations of a single form of demand. Experiment 2 used a within-subject design in which individuals foraged for varying forms of demand in addition to varying reward magnitudes. We leveraged the ecological and computational richness of foraging paradigms to provide new insights on a number of open questions [@Inzlicht2018]. First, we examined whether the discounting effects consistently observed in multi-alternative economic choice paradigms persisted when these options were presented serially, and whether the resulting foraging patterns depended on the demands imposed to obtain rewards. The existence of a well-defined optimal, reward-maximizing choice strategy helped us identify when demands boosted or detracted from the value of rewards (signified by tendencies to overharvest poor offers or underharvest advantageous offers, respectively). Second, we probed whether the perceived cost (or value) of demands was intrinsic or dependent on either task-specific or contextual characteristics. For instance, varying the level of exposure to multiple types of demands across experiments allowed us to assess whether cost evaluation was dependent on the availability of alternative avenues of action. Further, including cognitive demands and physical demands let us examine whether such effects might apply differently to distinct types of effort.

  Our results provided mixed support for our pre-registered hypotheses, which are detailed in https://osf.io/2rsgm/. For Experiment 1, we hypothesized that 1) decision makers would integrate reward and timing information in order to approximate optimal foraging strategies, as reflected by acceptance rates; 2) acceptance rates would be higher for trials that demanded physical effort and cognitive effort than for unfilled delay or effortless engagement (based on the apparent value of effort displayed by charitable donors, as described above); and 3) choices would be consistent within trials, reflected by few choice reversals, and acceptance rates for each type of demand would remain stable over time. Confirming our predictions, we found that individuals could approximate optimality by preferring higher rewards and shorter delays, regardless of the richness of the environment. We also found that people in the cognitive effort group accepted more trials than optimal, whereas physical effort and wait groups under-accepted, partially matching our predictions and highlighting that cognitive effort can add value to outcomes. Finally, acceptances (and differences among demands) were both consistent and stable within each cost condition, in line with expectations. For Experiment 2, we hypothesized that 1) acceptance rates would once again approximate optimality, and would show effects of reward magnitude similar to those seen in Experiment 1; 2) in contrast to Experiment 1, participants would gravitate towards unfilled delays to obtain rewards; and 3) that choices would be consistent and stable over time. As before, decision makers approached the optimal strategy for the foraging environment, resembling the pattern observed in Experiment 1. Contrary to our second and third hypotheses, we found that a preference for cognitive effort was present initially but disappeared over time as decision makers foraged for types of demands in addition to rewards.

  Computational modeling results suggested that the normative predictions of traditional foraging models were insufficient to explain our data. Instead, the data supported a model in which demands imposed a bias on the estimated opportunity cost of time, which converged as individuals gained experienced with interleaved types of demands. We hypothesize that this bias arises from modulations of subjective time. For instance, cognitive effort could have become appealing by virtue of compressing the perceived trial time, a perception that could be recalibrated with exposure to alternative demands. The subjective shortening of experienced durations during cognitive effort could be due to the ongoing recruitment of working memory (e.g. preventing individuals from estimating the elapsed time), or because the discrete events within the cognitive task subdivided the time interval. These results expand our understanding of the attractiveness of effort, suggesting potential reward-independent factors that could be leveraged to motivate effort engagement in diverse, everyday scenarios.

>>>>

  Evaluating whether to pursue a desirable outcome often entails assessing the time and effort it will require. In psychology and economics, demands for both time and effort have traditionally been identified as costs that detract from the net subjective value of rewards. Decision makers tend to prefer low-effort and immediate rewards over larger rewards that demand longer delays and higher effort [@Ainslie1975; @Hull1943; @Kool2010; @Kool2018; @Treadway2009; @Westbrook2013; @Shenhav2017; @Westbrook2015; @Frederick2002; @Walton2006]. However, people also sometimes appear to favor effortful courses of action in everyday decisions. Experimental and theoretical work has chiefly explained this phenomenon by suggesting a reciprocal relationship between effort and outcomes, whereby exerting effort boosts the perceived value of ensuing outcomes [@HernandezLallement2014; @Kacelnik2002; @Mochon2012], and positive outcomes imbue value into the preceding effortful behavior [i.e. learned industriousness, @Eisenberger1992]. These proposals raise questions about what features can make an effortful task attractive in its own right, and how contextual factors can alter its perceived costs [@Inzlicht2018]. Understanding what governs the intrinsic costs and rewards of effort could improve our ability to motivate completion of effortful daily activities such as schoolwork or physical exercise.

  Recent work has proposed that the subjective costs of delay and cognitive effort are modulated by the value of the alternative ways one's time and cognitive resources could otherwise be used, implying that decision makers track *opportunity costs* [@Fawcett2012; @Kurzban2013; @Otto2019]. Preferences for or against effortful courses of action could therefore depend on the presence of other types of demands and opportunities in the same choice environment [@Kool2018]. For example, exerting high effort for a reward might be aversive, but the same level of work might become appealing if the only alternative is the boredom of an equivalent period of passive waiting (for instance, a person might choose to take a cumbersome way home rather than wait for a delayed train, even if the resulting time of arrival is the same). Effects of context on effort preferences have been illustrated by experiments on charitable giving, in a which high-effort donation method (such as a run) attracted larger donations than a low-effort method (such as a picnic) when either was available alone (the “martyrdom effect”), but the high-effort event was disfavored when both options were offered as alternatives [@Olivola2013].

  **Standard models in neuroeconomics and behavioral economics formalize the effects of effort or delay on valuation in terms of discounting functions** [@Ainslie1975; @Frederick2002; @Green1994; @Kable2007; @kleinflugge2015; @Kool2010; @Westbrook2013]. Discounting functions appear to have a similar form for delay and effort [@Massar2015; @Prevost2010; @Seaman2018], although there is ongoing debate about how the shape of the discounting function might vary across types of demands [@Arulpragasam2018; @Biaaszek2017; @Chong2017; @Hartmann2013; @Klein-Flugge2016; @Kool2018; @Prevost2010] and the degree to which discount rates correlate across cost domains [@Seaman2018; @Westbrook2013]. A limitation of the discounting approach is that it does not directly probe how aspects of the decision context, such as opportunity costs, might influence cost evaluation, and constrains the understanding of behavioral demands to one in which such demands necessarily reduce the value of prospects.

  Single choice foraging paradigms, in which decision makers choose to accept or reject individually presented prospects, provide a natural way to investigate contextual influences on cost evaluation [@Constantino2015; @Garrett2020; @Krebs1977; @Mobbs2018; @Stephens1986]. Such paradigms require weighing a prospective reward against the perceived opportunity cost of time. Specifically, foragers must decide whether to spend their limited time by harvesting a currently available resource or by seeking out more profitable (or less costly) alternatives instead. This could involve, for example, deciding when to leave a patch of depleting richness **(e.g. when to quit a job that provides diminishing opportunities for growth)**, or choosing whether to pursue or forgo offers presented serially **(e.g. whether to accept your first job offer or wait for new potential ones)**. Here we focus on this latter type of foraging, traditionally called *prey selection* [@Krebs1977], as it provides a means to examine how choices to accept or reject individual prospects are influenced by contextual knowledge about other potentially obtainable alternatives in the same environment.

  Foraging has recently attracted high scientific interest due to its ecological validity, which is rooted in evolutionarily conserved choice behaviors [@Hayden2018; @Mobbs2018]. The prey selection foraging paradigm allows for the estimation of optimal, reward maximizing choice patterns on the basis of the opportunity cost of time [@Charnov1976; @Krebs1977]. The opportunity cost depends on the richness of the environment, which can be experimentally manipulated by balancing the amount of time it takes to harvest an offer (*handling time*) and to search for a new one (*travel time*). We can then interpret observed deviations from such optimal behavior, and examine how foraging behavior changes when the handling time is filled with cognitive or physical effort in comparison with an unfilled delay (thus disentangling effort from delay duration). In contrast to multi-alternative economic choice paradigms, which often treat choices as independent events, the sequential nature of foraging encourages individuals to consider both focal and global elements of the decision context. Extensive previous work on modeling foraging decisions provides computational tools to formalize relevant decision variables [@Stephens1986].

  Across two behavioral experiments, we tested the overarching hypothesis that preferences would vary depending on whether individuals faced a single type or multiple types of behavioral demands (i.e. cognitive effort, physical effort, and delay). Experiment 1 used a between-subject design in which each participant foraged for rewards of varying magnitudes that imposed varying durations of a single form of demand. Experiment 2 used a within-subject design in which individuals foraged for varying forms of demand in addition to varying reward magnitudes. First, we examined whether the discounting effects consistently observed in multi-alternative economic choice paradigms persisted when these options were presented serially, and whether the resulting foraging patterns depended on the demands imposed to obtain rewards. The existence of a well-defined optimal, reward-maximizing choice strategy helped us identify when demands boosted or detracted from the value of rewards (signified by tendencies to overharvest poor offers or underharvest advantageous offers, respectively). Second, we probed whether the perceived cost (or value) of demands was intrinsic or context-dependent. Varying the level of exposure to multiple types of demands across experiments allowed us to assess whether cost evaluation was dependent on the availability of alternative avenues of action. Further, including cognitive demands and physical demands let us examine whether such effects might apply differently to distinct types of effort.

  Our results provided mixed support for our pre-registered hypotheses, which are detailed in https://osf.io/2rsgm/registrations. For Experiment 1, we hypothesized that 1) decision makers would integrate reward and timing information in order to approximate optimal foraging strategies, as reflected by acceptance rates; 2) acceptance rates would be higher for trials that demanded physical effort and cognitive effort than for unfilled delay or effortless engagement (based on results from pilot data and previous findings from single-alternative choice paradigms; e.g., Olivola & Shafir, 2013. See also Eisenberger, 1992; Inzlicht et al., 2018); and 3) acceptance rates for each type of demand would remain stable over time, and participants would rarely exhibit within-trial reversals (accepting a trial but then quitting without receiving the reward). Consistent with our predictions, we found that individuals could approximate optimality by favoring higher rewards and shorter delays. We also found that people in the cognitive effort group accepted more trials than optimal, whereas physical effort and wait groups under-accepted, partially matching our predictions but raising the possibility that cognitive effort might add value to outcomes (or attenuate subjective temporal costs). Finally, acceptance rates (and differences in acceptance rates across demands) were stable within each cost condition, in line with expectations. These findings motivated the design of Experiment 2, in which effort and delay trials were intermixed so that rejecting a trial of one type could potentially lead to a trial of the other type. For Experiment 2, we hypothesized that 1) acceptance rates would once again approximate optimality, and would show effects of reward magnitude similar to those seen in Experiment 1; 2) in contrast to Experiment 1, participants would accept unfilled-delay trials at a higher rate than trials with either form of effort; and 3) that choices would be consistent and stable over time. As before, decision makers approached the optimal strategy for the foraging environment, resembling the pattern observed in Experiment 1. Contrary to our second and third hypotheses, we found that acceptance rates were initially higher for cognitive effort than for delay, but this apparent preference disappeared over time as decision makers foraged for types of demands in addition to rewards.

  Computational modeling results suggested that the normative predictions of traditional foraging models were insufficient to explain our data. Instead, the data supported a model in which demands imposed a bias on the estimated opportunity cost of time, which converged as individuals gained experience with interleaved types of demands. We hypothesize that this bias arises from modulations of subjective time. For instance, cognitive effort could have become appealing by virtue of compressing the perceived trial time, a perception that could be recalibrated with exposure to alternative demands. A subjective shortening of experienced durations during cognitive effort could be due to the ongoing recruitment of working memory (e.g. preventing individuals from estimating the elapsed time), or because the discrete events within the cognitive task subdivided the time interval. These results expand our understanding of the attractiveness of effort, suggesting potential reward-independent factors that could be leveraged to motivate effort engagement in diverse, everyday scenarios.
